---
title: "Reproducible Research Project 1"
author: "Ravinendra Pratap"
date: "10 December 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## R Markdown
# Reproducible Research: Peer Assessment 1
## Loading and preprocessing the data

```{r}
getwd()
setwd ("D:/LND/COURSERA_DATA_SCIENCE/COURSERA_05_Reproducible Research/WEEK2_05RR_Markdown_knitr/Assignment")
```
Loading and preprocessing the data

install.packages("ggplot2")
install.packages("dplyr")
install.packages("chron") 
```{r}
library(ggplot2)
library(dplyr)
library (chron)
```
##### 1. Load the data (i.e. read.csv())
##### Downloading zip file if it doesn't already exist in the workspace
```{r}
path <- getwd()
download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
              , destfile = paste(path, "dataFiles.zip", sep = "/"))
unzip(zipfile = "dataFiles.zip")
```
Clear the workspace load raw activity data
```{r}
rm(list=ls())
activity_raw <- read.csv("activity.csv", stringsAsFactors=FALSE)
```

##Process/transform the data suitable for analysis
### Transform the date attribute to an actual date format
```{r}
activity_raw$date <- as.POSIXct(activity_raw$date, format="%Y-%m-%d")
activity_raw <- data.frame(date=activity_raw$date, 
                           weekday=tolower(weekdays(activity_raw$date)), 
                           steps=activity_raw$steps, 
                           interval=activity_raw$interval)
```
Compute the day type (weekend or weekday)  
```{r}
activity_raw <- cbind(activity_raw, 
                      daytype=ifelse(activity_raw$weekday == "saturday" | 
                                       activity_raw$weekday == "sunday", "weekend", 
                                     "weekday"))

activity <- data.frame(date=activity_raw$date, 
                       weekday=activity_raw$weekday, 
                       daytype=activity_raw$daytype, 
                       interval=activity_raw$interval,
                       steps=activity_raw$steps)

rm(activity_raw)

```
Checking activity frame
```{r}
dim(activity)
head(activity)
str(activity)
summary(activity)
```
##### 1. Make a histogram of the total number of steps taken each day
```{r}
activity_total_steps <- with(activity, aggregate(steps, by = list(date), FUN = sum, na.rm = TRUE))
names(activity_total_steps) <- c("date", "steps")
hist(activity_total_steps$steps, main = "Total number of steps taken per day", xlab = "Total steps taken per day",  col = "lightblue", ylim = c(0,20), breaks = seq(0,25000, by=2500), labels=TRUE)
abline(v = mean(activity_total_steps$steps), lty = 1, lwd = 2, col = "red")
abline(v = median(activity_total_steps$steps), lty = 2, lwd = 2, col = "black")
legend(x = "topright", c("Mean", "Median"), col = c("red", "black"), 
       lty = c(1, 2), lwd = c(2, 2))

```
```{r}
##Mean
mean(activity_total_steps$steps)

##Median
median(activity_total_steps$steps)

summary(activity_total_steps$steps)
```

## What is the average daily activity pattern?   
## Excludes Missing Values"NA" using na.rm=TRUE   
```{r}
average_daily_activity <- aggregate(activity$steps, by=list(activity$interval), FUN=mean, na.rm=TRUE)
names(average_daily_activity) <- c("interval", "mean")

plot(average_daily_activity$interval, average_daily_activity$mean, type = "l", col="darkblue", lwd = 2, xlab="Interval", ylab="Average number of steps", main="Average number of steps per intervals")
average_daily_activity[which.max(average_daily_activity$mean), ]$interval
abline(v = average_daily_activity[which.max(average_daily_activity$mean), ]$interval, lty = 3, lwd = 2, col = "red")
legend(x = "topright", c("Max(Avg Daily Actvity Mean)"), col = c("red"),lty = c(3), cex=0.65)


## Max Average 
## average_daily_activity[which.max(average_daily_activity$mean), [1]

average_daily_activity[which.max(average_daily_activity$mean), ]$interval

## Maximum Average Number of Steps
## average_daily_activity[which.max(average_daily_activity$mean), ][2]

average_daily_activity[which.max(average_daily_activity$mean), ]$mean
```
# Split into two sets: complete and missing.
```{R}
activity.missing <- activity[is.na(activity$steps),]
activity.complete<-activity[complete.cases(activity),]

NA_count <- sum(is.na(activity$steps))
NA_pos <- which(is.na(activity$steps))
mean_vec <- rep(mean(activity$steps, na.rm=TRUE), times=length(NA_pos))
activity.complete[NA_pos, "steps"] <- mean_vec
head(activity.complete)
```

# Compute the total number of steps each day (NA values removed)
```{r}
sum_data <- aggregate(activity.complete$steps, by=list(activity.complete$date), FUN=sum)

## Rename the attributes
names(sum_data) <- c("date", "total")
```
# Compute the histogram of the total number of steps each day

```{r}
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="orange", 
     xlab="Total number of steps", 
     ylim=c(0, 30), 
     main="Histogram of the total number of steps taken each day\n(With missing data imputed\n NA Replaced by Mean value)")
```
```{r}
## Mean
 mean(sum_data$total)
## Median
median(sum_data$total)

## Clear the workspace
rm(sum_data)

## Load the lattice graphical library---
library(lattice)
```

# Compute the average number of steps taken, averaged across all daytype variable
```{r}
head(activity.complete)
activity.complete.daytype <- aggregate(steps ~ daytype+interval, data=activity.complete, FUN=mean)
head(activity.complete.daytype)
```

# Compute the time serie plot
```{r}

xyplot(steps ~ interval | daytype, activity.complete.daytype, 
       type="l", 
       lwd=1, 
       xlab="Interval", 
       ylab="Mean Number of steps", 
       layout=c(1,2))


## It seems that the weekday activities starts earlier than the weekends and weekday activities starts around 5-6am and weekend activities starts around 8am. 
## Another observation is that from 10am to 5pm in the weekends have higher activity levels than the weekdays.
